Tracing the Meta-Level: PyPy's Tracing JIT Compiler

### 2009-06-10 talk/icooolps2009/paper.tex
Enabling ACM categories.
Trying to get the copyright info right. No clue whether it is correct.
Now that we need 9 pages anyway, make the font non-small again. Add another reference.

\numberofauthors{4}
\author{
\alignauthor Carl Friedrich Bolz\\
       \affaddr{University of Düsseldorf}\\
       \affaddr{STUPS Group}\\
       \affaddr{Germany}\\
       \email{cfbolz@gmx.de}
\alignauthor Antonio Cuni\\
       \affaddr{University of Genova}\\
       \affaddr{DISI}\\
       \affaddr{Italy}\\
       \email{cuni@disi.unige.it}
\and
\alignauthor Maciej Fijalkowski\\
       \affaddr{merlinux GmbH}\\
       \email{fijal@merlinux.eu}
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
}
\maketitle

%\category{D.3.4}{Programming Languages}{Processors}[code generation,
%interpreters, run-time environments]
%\category{F.3.2}{Logics and Meanings of Programs}{Semantics of Programming
%Languages}[program analysis]

@@

\begin{figure}
{\small
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result

@@

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot interpreter loop
is the bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption made by the tracing JIT -- that several iterations of a hot loop
take the same or similar code paths -- is wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{\small
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

\begin{figure}[h]
\label{fig:bench-example}
{\small
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &PyPy compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &PyPy comp'd to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.2 &4.96 $\pm$ 0.05 &4.73\\
4 &CPython 2.5.2 + Psyco 1.6 &1.51 $\pm$ 0.05 &15.57\\\hline
\end{tabular}

@@

There are quite a number of approaches that try to minimally enhance
interpreters to generate code at runtime without actually writing a native
compiler by hand. The goal of these is to get rid of dispatch overhead of
typical interpreters while retaining ease of implementation. Piumarta and
Riccardi \cite{piumarta_optimizing_1998} propose to copy fragments of the
interpreter together for commonly occurring bytecode sequences to reduce
dispatch overhead. However, dispatching is still needed to jump between such
sequences and also when non-copyable bytecodes occur. Ertl and Gregg
\cite{ertl_retargeting_2004} go further and stitch together the concatenated
sequences by patching the copied machine code. Thus they get rid of all dispatch
overhead. Both techniques can speed up interpreters which large dispatch
overhead a lot. However they will help less if the bytecodes themselves do a
lot of work (as is the case with Python) and the dispatch overhead is lower. On
the other hand, our technique can do a better job by tracing inside the
implementation of those bytecode and inlining common paths.

>>>>

\numberofauthors{4}
\author{
\alignauthor Carl Friedrich Bolz\\
       \affaddr{University of Düsseldorf}\\
       \affaddr{STUPS Group}\\
       \affaddr{Germany}\\
       \email{cfbolz@gmx.de}
\alignauthor Antonio Cuni\\
       \affaddr{University of Genova}\\
       \affaddr{DISI}\\
       \affaddr{Italy}\\
       \email{cuni@disi.unige.it}
\and
\alignauthor Maciej Fijalkowski\\
       \affaddr{merlinux GmbH}\\
       \email{fijal@merlinux.eu}
\alignauthor Armin Rigo\\
       \email{arigo@tunes.org}
}
\conferenceinfo{ICOOOLPS}{'09 Genova, Italy}
\CopyrightYear{2009}
\crdata{978-1-60558-541-3/09/07}

\maketitle

\category{D.3.4}{Programming Languages}{Processors}[code generation,
incremental compilers, interpreters, run-time environments]

@@

\begin{figure}
{
\begin{verbatim}
def f(a, b):
    if b % 46 == 41:
        return a - b
    else:
        return a + b
def strange_sum(n):
    result = 0
    while n >= 0:
        result = f(result, n)
        n -= 1
    return result

@@

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The most important hot interpreter loop
is the bytecode dispatch loop (for many simple
interpreters it is also the only hot loop).  Tracing one iteration of this
loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption made by the tracing JIT -- that several iterations of a hot loop
take the same or similar code paths -- is wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.
\begin{figure}
\input{code/tlr-paper.py}
\caption{A very simple bytecode interpreter with registers and an accumulator.}
\label{fig:tlr-basic}
\end{figure}
\begin{figure}
{
\begin{verbatim}
    MOV_A_R     0   # i = a
    MOV_A_R     1   # copy of 'a'

@@

\begin{figure}[h]
\label{fig:bench-example}
{
\begin{verbatim}
def f(a):
    t = (1, 2, 3)
    i = 0
    while i < a:
        t = (t[1], t[2], t[0])
        i += t[0]
    return i
\end{verbatim}
}
\noindent
\begin{tabular}{|l|l|r|r|}
\hline
& &Time (s) &speedup\\
\hline
1 &PyPy compiled to C, no JIT &23.44 $\pm$ 0.07 &1.00\\
2 &PyPy comp'd to C, with JIT &3.58 $\pm$ 0.05 &6.54\\
3 &CPython 2.5.2 &4.96 $\pm$ 0.05 &4.73\\
4 &CPython 2.5.2 + Psyco 1.6 &1.51 $\pm$ 0.05 &15.57\\\hline
\end{tabular}

@@

There are quite a number of approaches that try to minimally enhance
interpreters to generate code at runtime without actually writing a native
compiler by hand. The goal of these is to get rid of dispatch overhead of
typical interpreters while retaining ease of implementation. Piumarta and
Riccardi \cite{piumarta_optimizing_1998} propose to copy fragments of the
interpreter together for commonly occurring bytecode sequences to reduce
dispatch overhead. However, dispatching is still needed to jump between such
sequences and also when non-copyable bytecodes occur. Ertl and Gregg
\cite{ertl_retargeting_2004} go further and stitch together the concatenated
sequences by patching the copied machine code. Thus they get rid of all dispatch
overhead. Both techniques can speed up interpreters which large dispatch
overhead a lot. However they will help less if the bytecodes themselves do a
lot of work (as is the case with Python
\cite{stefan_brunthaler_virtual-machine_2009}) and the dispatch overhead is lower. On
the other hand, our technique can do a better job by tracing inside the
implementation of those bytecode and inlining common paths.
